---
title: "MSDS6306CaseStudy"
author: "Priyanka Tilak"
date: "March 19, 2017"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

## Introduction to the Project
This document describes the Gross Domestic Product data and Educational data for 190 countries across the world. We are analyzing GDP of the countries, different income groups of these countries and their contribution in million US dollars of Gross domestic product in each country.

We have downloaded 2 data files, 1 for GDP data and Countries educational data from the following web sources:

GDP of 190 countries: https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv 

Educational statistics of Different countries: https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv

#### Information about raw data files:

* Both the raw data files are in Comma separated format (csv)

* GDP file has NAs, missing heading, missing rows, content which is not very informative

* Education information data files has NAs

* Filtering required to perform analysis on these files

#### Objectives:

* Merge the data based on the country short code. How many of the IDs match?

* Sort the data frame in ascending order by GDP (so United States is last). What is the 13th
  country in the resulting data frame? 
  
* What are the average GDP rankings for the "High income: OECD" and "High income:
  nonOECD" groups? 
  
* Show the distribution of GDP value for all the countries and color plots by income group. Use
  ggplot2 to create your plot.
  
* Provide summary statistics of GDP by income groups.

* Cut the GDP ranking into 5 separate quantile groups. Make a table versus Income.Group.
  How many countries are Lower middle income but among the 38 nations with highest
  GDP?

To begin with we will first set up a working directory.
```{r warning=FALSE}
#Set the present directory as working directory
setwd(".")
#view present working directory
getwd()
```

Setting up the working directory and downloading data files from web resources 
```{r warning=FALSE}
#Downloading GDP file from the WEB URL
download.file("https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv",destfile = ".\\Data\\GDP.csv", mode = "w")

#Downloading Edstats_country file from the WEB URL
download.file("https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv",destfile = ".\\Data\\Edstats_Country.csv", mode= "w")

#Libraries Required
library(downloader)
library(plyr)
library(dplyr)
library(ggplot2)
library(sqldf)

```

#### Problems with Raw Data files

Raw data files required cleaning in order to move further for analysis and merging

* GDP data file has empty rows

* GDP data file has empty column

* GDP data file needs to have fix for column headings

* GDP data file has data for 190 countries and other geographical regions which are not country

* GDP for countries has NAs and missing value

* Education data for country has many columns which are not required for the analysis or objectives

* Education data for countries has NA's and missing value


## Data Wrangling 

To clean the raw data files in order to use it for merging and for analysis, Data Wrangling step is performed
```{r warning=FALSE}
#read the Downloaded files in objects
GDP<-read.csv(".\\Data\\GDP.csv",sep=",",header=F, skip= 5)
Edstats_Country<-read.csv("C:\\Personal\\SMU\\SAS\\R\\R_REPO\\MSDS6306CaseStudy1\\Data\\Edstats_Country.csv",sep=",",header=T)

library(plyr)
#GDP data file cleanup\n
#giving new column names and ignoring empty rows for GDP table
colnames(GDP) <- c("CountryCode","Rank","Empty","CountryName","GDPA" )
keep_columns<-c("CountryCode","Rank","CountryName","GDPA")
GDP<-GDP[keep_columns]

# Calculating missing values and NAs before cleaning up GDP data
sum(is.na(GDP$CountryCode) | GDP$CountryCode=="")
sum(is.na(GDP$Rank) | GDP$Rank=="")
sum(is.na(GDP$GDPA) | GDP$GDPA =="")

#Cleaning GDP file data
GDP<-GDP[!(is.na(GDP$CountryCode) | GDP$CountryCode=="" | is.na(GDP$GDPA) | GDP$GDPA=="" | is.na(GDP$Rank) | GDP$Rank==""),]
#view first 5 rows of the dataframe
head(GDP)

#Cleaning Edstats_Country data file
colnames(Edstats_Country)[1] <- "CountryCode"
colnames(Edstats_Country)[2] <- "LongName"
colnames(Edstats_Country)[3] <- "IncomeGroup"

# Calculating missing values and NAs before cleaning up Edstats_Country data
sum(is.na(Edstats_Country$CountryCode) | Edstats_Country$CountryCode=="")
sum(is.na(Edstats_Country$Income.Group) | Edstats_Country$Income.Group=="")

library(sqldf)
Edstats_Country<-sqldf("select Edstats_Country.CountryCode,Edstats_Country.LongName, Edstats_Country.IncomeGroup from Edstats_Country ")
#view first 5 rows of the dataframe
head(Edstats_Country)
```

## Analysis

To perform the analysis on GDP of the countries and their income groups, both cleaned data files "GDP"" and Edstats_Country need to be merged

### Objective 1: Merge the data based on the country short code. How many of the IDs match?

Following is the Merged Table and its structure, which is merged table of GDP and Edstats_Country

```{r warning=FALSE}
#merging 2 tables by CountryCode in to new data frame Merged_Table
Merged_Table<-merge(GDP,Edstats_Country, by="CountryCode")
#viewving first five records of the merged data frame "Merged_Table" 
head(Merged_Table)

#removing blank and "," values in GDPA and Rank column and making it as numeric column
Merged_Table$GDPA <- as.numeric(gsub(",","", Merged_Table$GDPA))
Merged_Table$Rank <- as.numeric(gsub(",","", Merged_Table$Rank))
#structure of the Merged_Table
str(Merged_Table)
```

### Objective 2: Sort the data frame in ascending order by GDP (so United States is last). What is the 13th country in the resulting data frame?

Following is the result of merging and sorting to get United State as a last country in the list, and 13th country in the list 
```{r warning=FALSE}
sortedMergeTable <- Merged_Table[ order(Merged_Table$GDPA), ]
#sorted merged table, "sortedMergeTable""
head(sortedMergeTable)
#to create output file
sortedMergeTable<-sortedMergeTable[(!(is.na(sortedMergeTable$IncomeGroup) |sortedMergeTable$IncomeGroup =="")),]
write.csv(sortedMergeTable, file = ".\\Data\\sortedMergeTable.csv")
head(sortedMergeTable)
#to view 13th country
sortedMergeTable[13,3]

```

### Objective 3: What are the average GDP rankings for the High income: OECD and High income: nonOECD groups?

Average Ranking for the different Income group and their respective mean in column 'X'
```{r warning=FALSE}
library(dplyr)
AvgRanking<-aggregate(sortedMergeTable$Rank,list(sortedMergeTable$IncomeGroup),mean)
print(AvgRanking)
```

### Objective 4: Show the distribution of GDP value for all the countries and color plots by income group. Use ggplot2 to create your plot.

Bar chart representation of GDP values of the countries and their Income group Contributions
```{r warning=FALSE}
library(ggplot2)
library(scales)
ggplot(sortedMergeTable,aes(x=IncomeGroup, y=GDPA,group_by(sortedMergeTable$IncomeGroup)))+ geom_bar(aes(fill=sortedMergeTable$IncomeGroup), stat = "identity") + scale_fill_brewer(palette = "Spectral") + theme( axis.text.x=element_text(angle=90)) +scale_y_continuous(labels = comma)
```

### Objective 5: Provide summary statistics of GDP by income groups

Summary Statistics of the complete merged data set for GDP
```{r warning=FALSE}
library('dplyr')
#Provide summary statistics of GDP by income groups.
summary(sortedMergeTable$GDPA, by=sortedMergeTable$IncomeGroup)
```

Summary Statistics of the complete merged data set for GDP for Lower Income group
```{r warning=FALSE}
#Statistics for Lower Income group
lowerIncome<-sortedMergeTable[sortedMergeTable$IncomeGroup=="Low income",]
print(summary(lowerIncome$GDPA))
```

Summary Statistics of the complete merged data set for GDP for Lower Middle Income group
```{r warning=FALSE}
#Statistics for Lower Middle Income group
lowerMiddleIncome<-sortedMergeTable[sortedMergeTable$IncomeGroup=="Lower middle income",]
print(summary(lowerMiddleIncome$GDPA))
```

Summary Statistics of the complete merged data set for GDP for Upper Middle Income group
```{r warning=FALSE}
#Statistics for Upper Middle Income group
upperMiddleIncome<-sortedMergeTable[sortedMergeTable$IncomeGroup=="Upper middle income",]
print(summary(upperMiddleIncome$GDPA))
```

Summary Statistics of the complete merged data set for GDP for High Income group
```{r warning=FALSE}
#Statistics for High Income group
highIncomeOECD<-sortedMergeTable[sortedMergeTable$IncomeGroup=="High income: OECD",]
print(summary(highIncomeOECD$GDPA))
```

Summary Statistics of the complete merged data set for GDP for High Income group
```{r warning=FALSE}
#Statistics for High Income Income group
highIncomeNonOECD<-sortedMergeTable[sortedMergeTable$IncomeGroup=="High income: nonOECD",]
summary(highIncomeNonOECD$GDPA)

```

#### Objective 6: Cut the GDP ranking into 5 separate quantile groups. Make a table versus Income.Group. How many countries are Lower middle income but among the 38 nations with highest GDP?
```{r warning=FALSE}
quantileGrps <-quantile(sortedMergeTable$Rank, probs=seq(0,1,0.2),na.rm=TRUE)
print(quantileGrps)
gdpQuantile <- cut(sortedMergeTable$Rank, breaks = quantileGrps)
#print(gdpQuantile)
sortedMergeTable$RankGroup <- gdpQuantile
tableIncGrpVsRnkGrp <- table(sortedMergeTable$IncomeGroup, sortedMergeTable$RankGroup)
tableIncGrpVsRnkGrp
```

Countries count which are lower middle income but among the 38 nations with the highest GDP
```{r warning=FALSE}
lower_mid_inc_count <-sqldf("select count(*) from sortedMergeTable where IncomeGroup=='Lower middle income' and  Rank <39")
print("Countries count which are lower middle income but among the 38 nations with the highest GDP")
print(lower_mid_inc_count)
```

## Conclusion

* According to the plot, the High Income OECD group is leading the total GDP as compared to all the other groups, followed surprisingly by the Lower middle income group. This indicates that the Lower Middle income group contribute significantly to the overall GDP of the world. 

* Looking at the statistical summary across various groups, we can see that the median income for High Income OECD group, comprising of only 23 countries, is atleast 10 times the median income of the other groups. 

* According to the plot, GDP of countries in the Lower middle income and Upper middle income are very comparable. 

